---
title: "The Long And Winding Road"
description: How much road footage does Franklin have.
categories: [Roads, Capital Improvements]
format: html
date: 2/21/2026
date-modified: 2/21/26
execute:
  echo: false
  warning: false
---
# Introduction
Talk to any person in town and they will tell you that the roads need to be repaired. THe state of Franklin's roads represent years of underinvestment, but how big is the problem we are talaking about? The best place to look for the answer to this question is Franklin's [Capital Improvement Plan (CIP)](https://www.franklinnh.gov/capital-improvement-plan-cip-committee/pages/capital-improvements-plan-cip-documents). The most recent plan drafted in 2024 identifies 21 roads that are in need of repair. This includes places like Davis street, Prosepct street and Cheney Hill. Note that the CIP identifies other capital needs as well; for example  Fire trucks and building repairs. 


# Draft CIP
This draft CIP calls for 4.6 million dollars of CIP funding in 2025, 5.9 million dollars in 2026, 5.9 million dollars in 2027 and increases from there. The city did not fund the CIP-identified projects fully in 2025, and likely will not be able to in 2026. The need for these repairs has not gone away, rather they are pushed back. THe committee of 9 who writes the CIP is mandated to meet at least once a year by the charter; I hope that they revist and engage with this issue. 

```{r read in data}
library(tidyverse)
library(osmdata)
library(sf)
library(tigris)
library(tmap)
tmap_mode("view")
options(tigris_use_cache = TRUE)
# set different Overpass server
set_overpass_url("https://overpass.private.coffee/api/interpreter")

franklin_city_limits <- places(state = "NH") |>
  filter(NAME == "Franklin") |>
  st_transform(crs = "EPSG:4326")

st_area(franklin_city_limits) %>% units::set_units(mi^2)

frank_bb <- st_bbox(franklin_city_limits)


x <- opq(bbox = frank_bb) |>
  add_osm_feature(
    key = "highway",
    value = c(
      "primary",
      "secondary",
      "tertiary",
      "motorway",
      "trunk",
      "unclassified",
      "residential"
    )
  ) |>
  osmdata_sf()

roads <- x$osm_lines



``` 



## Roads
The city of Franklin is fairly small, only 29.1 square miles. [Open Street Map](https://www.openstreetmap.org/#map=8/43.461/-71.691) (OSM) is an open-source, community-editable map. It's kind of like wikipedia for place information and is used by both private and public entities as a source of geopgraphical information.

 Pulling data from OSM we can get a sense for how many roads are in the city and what kinds of roads they are (paved, dirt etc). I'm focusing on surface type for this blog as it is speaks more to the maintenance need. Future posts may look at speed limits for traffic safety, walking paths through the city for non-car based travel or other interesting transit questions. 



```{r number and type}



frank_roads <- roads[franklin_city_limits, op = st_intersects] %>% select(highway, surface)


frank_roads |>
  tm_shape() +
  tm_lines(col = "highway", lwd = 10)

frank_roads %>%
    mutate(length = st_length(geometry)%>% units::set_units(mi) %>% as.numeric() %>% round(2)) %>% st_drop_geometry() %>% 
    group_by(highway) %>%
    summarise(n = n(), 
    tot_length = sum(length)) %>%
        gt() %>%
        cols_label(
      .list = c(
        n = "Number of Roads",
        tot_length = "Total Road Length (miles)")) %>% 
            grand_summary_rows(fns = list(label = "Total", fn = "sum"), 
            columns = where(is.numeric))


frank_roads |>
  tm_shape() +
  tm_lines(col = "surface", lwd = 10)



```



The city has a 

OSM Road length
OSM road type
direct people to the CIP
