---
title: "How Much Property does the city own?"
description: "Using the axis GIS to figure out where and how much property the city owns"
author: "S Jacobson "
date: "2025-12-27"
execute:
  echo: false
  warning: false
categories:
  - analysis
---
A state priorty of the city manager is to return city property to the tax rolls. Most recently this has come before the council working with the snowmobile club. Because the city does not pay taxes on land that owns, City-owned land represents lost revenue. This blog post will look at the location and tax value of city owned land. 


```{r load libs}
library(tidyverse)
library(sf)
library(tmap)
library(gt)
tmap_mode("view")

```

```{r read in data}

parcel_polys_fp<- here::here(".data/20251116_geo_dat/parcel_polys.shp")
parcel_polys<- st_read(dsn = parcel_polys_fp, quiet = T)
zone_fp<- here::here(".data/20251116_geo_dat/zone_polys.shp")
zone_polys<- st_read(zone_fp, quiet = T)

dat<- read_csv(here::here(".data/outputs/prop_data_with_zone.csv"))

city_data<- dat %>% filter(OwnerName == "FRANKLIN, CITY OF" ) 

```


Let's look at city property. The city owns a series of parcels around within the municipality. Over time the city acquires property. This land may be conservation land, it may have been taken for tax purposes or it may be used for other city purposes. An important priority stated by our City manager is returning property to the tax rolls. 

```{r map it}

city_data%>%
  summarize(n = n(),
  av_size= mean(ACRES, na.rm= T), 
  av_tax_val = mean(TOTTXVAL, na.rm =T), 
  med_size= median(ACRES, na.rm= T), 
  med_tax_val = median(TOTTXVAL, na.rm =T)) %>%
      gt() |>
    cols_label(
      .list = c(
        n = "Number of Properties",
        av_size = "Average Property Size (Acres)", 
        av_tax_val = "Average Tax Value ($)", 
        med_size = "Median size (acres)", 
        med_tax_val = "Median Tax Value ($)")) %>%
        fmt_number(c(av_size, med_size)) %>%
    fmt_currency(c(av_tax_val, med_tax_val)) %>%
     opt_interactive(use_pagination = FALSE)

##looks like this includes roads###
city_shp<- city_data %>%
  left_join(parcel_polys) %>% st_as_sf()

tm_shape(city_shp) + 
  tm_polygons() + 
  tm_basemap(leaflet::providers$OpenStreetMap) + 
  tm_title("Franklin City Property")

```

This table shows that the city owns about 149 different parcels throughout the city! That's a lot of parcels. I've included the average and median size and value of the parcels as well. Looking at the map above you can see there is a wide range in what that city owns. Some (like 266 flag hole road) are large and represent 366 acres of land. Others (like some of the properties on East bow Street) are small (0.08) acres.

Another way to look at this is to look at the zone that these parcels fall in - if returned to the rolls, how much is conservation, light industrial etc? 

```{r }
city_data%>% group_by(NAME) %>% 
  replace_na(list(NAME = "Missing Zone")) %>% 
  summarize(n = n(),
  av_size= mean(ACRES, na.rm= T), 
  av_tax_val = mean(TOTTXVAL, na.rm =T), 
  med_size= median(ACRES, na.rm= T), 
  med_tax_val = median(TOTTXVAL, na.rm =T)) %>%
      gt() |>
    cols_label(
      .list = c(
        n = "Number of Properties",
        NAME = "Zoned Category",
        av_size = "Average Property Size (Acres)", 
        av_tax_val = "Average Tax Value ($)", 
        med_size = "Median size (acres)", 
        med_tax_val = "Median Tax Value ($)")) %>%
        fmt_number(c(av_size, med_size)) %>%
    fmt_currency(c(av_tax_val, med_tax_val)) %>%
     opt_interactive(use_pagination = FALSE)



tm_shape(city_shp) + 
  tm_polygons("NAME",  fill.scale = tm_scale_categorical(values.repeat = F)) +  
  tm_basemap(leaflet::providers$OpenStreetMap) + 
  tm_title("Franklin City Property")

```

If you look closely at the above maps, you'll see that this datset includes things like roads and slivers 